Efficiently Handling Guards in the Low Level Design of RPython's tracing JIT

### 2012-08-03 talk/vmil2012/paper.tex
move figure to appendix

%\begin{figure*}
%    \include{figures/ops_count_table}
%    \caption{Relative numbers of operations in the traces generated for
%    different benchmarks}
%    \label{fig:ops_count}
%\end{figure*}
Although there are several publications about tracing just-in-time compilers,
to our knowledge, there are none that describe deoptimization and the use and
implementation of guards in this context.

@@

After optimization the resulting trace is handed to the backend to be compiled
to machine code. The compilation phase consists of two passes over the lists of
instructions, a backwards pass to calculate live ranges of IR-level variables
and a forward one to emit the instructions. During the forward pass IR-level
variables are assigned to registers and stack locations by the register
allocator according to the requirements of the to be emitted instructions.
Eviction/spilling is performed based on the live range information collected in
the first pass. Each IR instruction is transformed into one or more machine
level instructions that implement the required semantics, operations withouth
side effects whose result is not used are not emitted. Guards instructions are
transformed into fast checks at the machine code level that verify the
corresponding condition.  In cases the value being checked by the guard is not
used anywhere else the guard and the operation producing the value can merged,
reducing even more the overhead of the guard. Figure \ref{fig:trace-compiled}
shows how an \texttt{int\_eq} operation followed by a guard that checks the
result of the operation are compiled to pseudo-assembler if the operation and
the guard are compiled separated or if they are merged.

@@

The following analysis is based on a selection of benchmarks taken from the set
of benchmarks used to measure the performance of PyPy as can be seen
on.\footnote{http://speed.pypy.org/} The benchmarks were taken from the PyPy benchmarks
repository using revision
\texttt{ff7b35837d0f}.\footnote{https://bitbucket.org/pypy/benchmarks/src/ff7b35837d0f}
The benchmarks were run on a version of PyPy based on the
tag~\texttt{release-1.9} and patched to collect additional data about the
guards in the machine code
backends.\footnote{https://bitbucket.org/pypy/pypy/src/release-1.9} All
benchmark data was collected on a MacBook Pro 64 bit running Max OS 10.8 with
the loop unrolling optimization disabled.\footnote{Since loop unrolling
duplicates the body of loops it would no longer be possible to meaningfully
compare the number of operations before and after optimization. Loop unrolling
is most effective for numeric kernels, so the benchmarks presented here are not
affected much by its absence.}

Figure~\ref{fig:benchmarks} shows the total number of operations that are
recorded during tracing for each of the benchmarks on what percentage of these
are guards. Figure~\ref{fig:benchmarks} also shows the number of operations left
after performing the different trace optimizations done by the trace optimizer,
such as xxx. The last columns show the overall optimization rate and the
optimization rate specific for guard operations, showing what percentage of the
operations was removed during the optimizations phase.

@@

Figure~\ref{fig:backend_data} shows
the total memory consumption of the code and of the data generated by the machine code
backend for the different benchmarks mentioned above. Meaning the operations
left after optimization take the space shown in Figure~\ref{fig:backend_data}
after being compiled. Also the additional data stored for the guards to be used
in case of a bailout and attaching a bridge.
\begin{figure*}
    \include{figures/backend_table}
    \caption{Total size of generated machine code and guard data}
    \label{fig:backend_data}
\end{figure*}

@@

\section*{Acknowledgements}
\bibliographystyle{abbrv}
\bibliography{zotero,paper}
\listoftodos
\end{document}

>>>>

Although there are several publications about tracing just-in-time compilers,
to our knowledge, there are none that describe deoptimization and the use and
implementation of guards in this context.

@@

After optimization the resulting trace is handed to the over platform specific
backend to be compiled to machine code. The compilation phase consists of two
passes over the lists of instructions, a backwards pass to calculate live
ranges of IR-level variables and a forward one to emit the instructions. During
the forward pass IR-level variables are assigned to registers and stack
locations by the register allocator according to the requirements of the to be
emitted instructions.  Eviction/spilling is performed based on the live range
information collected in the first pass. Each IR instruction is transformed
into one or more machine level instructions that implement the required
semantics, operations withouth side effects whose result is not used are not
emitted. Guards instructions are transformed into fast checks at the machine
code level that verify the corresponding condition.  In cases the value being
checked by the guard is not used anywhere else the guard and the operation
producing the value can merged, reducing even more the overhead of the guard.
Figure \ref{fig:trace-compiled} shows how an \texttt{int\_eq} operation
followed by a guard that checks the result of the operation are compiled to
pseudo-assembler if the operation and the guard are compiled separated or if
they are merged.

@@

The results presented in this section are based on numbers gathered by running
a subset of the standard PyPy benchmarks. The PyPy benchmarks are used to
measure the performance of PyPy and are composed of a series of
micro-benchmarks and larger programs.\footnote{http://speed.pypy.org/} The
benchmarks were taken from the PyPy benchmarks repository using revision
\texttt{ff7b35837d0f}.\footnote{https://bitbucket.org/pypy/benchmarks/src/ff7b35837d0f}
The benchmarks were run on a version of PyPy based on the
tag~\texttt{0b77afaafdd0} and patched to collect additional data about the
guards in the machine code
backends.\footnote{https://bitbucket.org/pypy/pypy/src/0b77afaafdd0} All
benchmark data was collected on a MacBook Pro 64 bit running Max OS 10.8 with
the loop unrolling optimization disabled.\footnote{Since loop unrolling
duplicates the body of loops it would no longer be possible to meaningfully
compare the number of operations before and after optimization. Loop unrolling
is most effective for numeric kernels, so the benchmarks presented here are not
affected much by its absence.}

Figure~\ref{fig:benchmarks} shows the total number of operations that are
recorded during tracing for each of the benchmarks and what percentage of these
are guards. Figure~\ref{fig:benchmarks} also shows the number of operations
left after performing the different trace optimizations done by the trace
optimizer, such as xxx. The last columns show the overall optimization rate and
the optimization rate specific for guard operations, showing what percentage of
the operations were removed during the optimizations phase.
Figure~\ref{fig:benchmarks} shows that as can also be seen on
Figure~\ref{fig:guard_percent} the optimization rate for guards is on par with
the average optimization rate for all operations in a trace. After optimization
the amount of guards left in the trace still represents about 15.18\% to
20.22\% of the operation, a bit less than before the optimization where guards
represented between 15.85\% and 22.48\% of the operations. After performing the
optimizations the most common operations are those that are difficult or
impossible to optimize, such as JIT internal operations and different types of
calls. These account for 14.53\% to 18.84\% of the operations before and for
28.69\% to 46.60\% of the operations after optimization. These numbers show
that about one fifth of the operations, making guards one of the most common
operations, that are compiled are guards and have associated with them the
high- and low-level datastructes that are reconstruct the state.

@@

The overhead that is incurred by the JIT to manage the \texttt{resume data},
the \texttt{low-level resume data} and the generated machine code is shown in
Figure~\ref{fig:backend_data}. It shows the total memory consumption of the
code and of the data generated by the machine code backend for the different
benchmarks mentioned above. The size of the machine code is composed of the
size of the compiled operations, the trampolines generated for the guards and a
set of support functions that are generated when the JIT starts and are shared
by all compiled traces. The size of the \texttt{low-level resume data} is the
size of the registers and stack to IR-level variable mappings and finally the
size of the \texttt{resume data} is an approximation of the size of the
compressed high-level resume data. While the \texttt{low-level resume data} has
a size of about 15\% to 20\% of the generated instructions the \texttt{resume
data} is even in the compressed form larger than the generated machine code.

Tracing JITs compilers only compile a subset of the executed program so the
amount of generated machine code will be smaller than for function based JITs.
At the same time there is a several times larger overhead for keeping the
resume information for the guards. The generated machine code accounts for
20.21\% to 37.97\% of the size required for storing the different kinds of
resume data.

\begin{figure*}
    \include{figures/backend_table}
    \caption{Total size of generated machine code and guard data}
    \label{fig:backend_data}
\end{figure*}

@@

\section*{Acknowledgements}
\section*{Appendix}
\begin{figure*}
    \include{figures/ops_count_table}
    \caption{Relative numbers of operations in the traces generated for
    different benchmarks}
    \label{fig:ops_count}
\end{figure*}
\bibliographystyle{abbrv}
\bibliography{zotero,paper}
\listoftodos
\end{document}
